\documentclass[11pt]{article}
  \usepackage{pslatex}
  \usepackage{amsmath}
  \advance\textwidth by2cm
  \advance\oddsidemargin by-1cm
  \advance\textheight by3cm
  \headheight 0pt\headsep 0pt

\begin{document}
  \thispagestyle{empty}

\begin{center}
  \Large\sf Advanced Algorithms, Fall 2012\\
  
  \medskip
  \large\sf Prof. Bernard Moret\\

  \medskip\bigskip
  \Large\bf Homework Assignment \#5\\

  \smallskip
  \normalsize\rm due Sunday night, Oct.\ 28
\end{center}

\bigskip\rm\noindent
\textit{Write your solutions in LaTeX using the template provided on the
Moodle and web sites and upload your PDF file through Moodle by 4:00 am
Monday morning, Oct.\ 29.}


\bigskip\rm\noindent
\emph{Question 1.~(Greedy Algorithms)}\\
You are given an undirected connected graph $G=(V, E)$ such that $|E| = |V| +
5$ (i.e., this is a very sparse graph).  Each edge of $G$ has a distinct cost.
Design a linear-time algorithm to find the Minimum Spanning Tree of $G$, and
prove the correctness of your algorithm.

\bigskip\rm\noindent
\emph{Question 2.~(Greedy Algorithms)}\\
Given $n$ positive integers, $a_1, a_2,\cdots, a_n$, design an polynomial-time
algorithm to determine whether there exists an undirected simple graph
(self loops and multi-edges are not allowed) with $n$ nodes such that
the degrees of the $n$ nodes correspond to the given $n$ integers; prove the
correctness of your algorithm.

\bigskip\rm\noindent
\emph{Question 3.~(Greedy Algorithms)}\\
This problem addresses optimal file allocation for computer networks.  You are
given a completely connected network of nodes, a set of files to be allocated
among the nodes, and a sequence of retrieval and updating requests (the entire
sequence is known in advance).  A retrieval or updating request is a triple,
consisting of the node initiating the request, the file involved, and the
number of bytes to be transferred.
  
An allocation scheme is an assignment of each file to one or more nodes.
Having multiple copies of a file is advantageous in retrieval: the cost of a
retrieval is zero when the file is held locally, but equals the number of bytes
to be transferred when the file must be accessed remotely.  On the other hand,
multiple copies of a file increase the cost of an updating operation, because
each copy must be updated and thus the number of bytes needed for updating is
multiplied by the number of remote copies.  Multiple copies are also desirable
for reasons of reliability, but only up to a point, since relatively few nodes
are expected to fail.  The gain in reliability is a function of the number of
copies and obeys the law of diminishing returns: each additional copy of a file
gives a smaller gain than the previous copy.  The cost of an allocation scheme
is the cost of the given sequences of retrieval and updating requests minus the
gain in reliability.

Develop a greedy algorithm that constructs the optimal allocation scheme
and prove its correctness.

\bigskip\rm\noindent
\emph{Question 4.~(Matching)}\\
We are given a $n\times n$ 0-1 matrix $M$.  We can swap two columns of $M$, or
two rows of $M$.  We say that $M$ can be diagonalized if we can perform a
sequence of (row and/or column) swaps to make all the diagonal items of $M$
equal to 1.  Design an algorithm to decide whether a given binary matrix $M$
can be diagonalized and prove the correctness of your algorithm.
\end{document}
