\documentclass[11pt]{article}
  \usepackage{pslatex}
  \usepackage{amsmath}
  \advance\textwidth by2cm
  \advance\oddsidemargin by-1cm
  \advance\textheight by3cm
  \headheight 0pt\headsep 0pt

\begin{document}
  \thispagestyle{empty}

\begin{center}
  \Large\sf Advanced Algorithms, Fall 2012\\
  
  \medskip
  \large\sf Prof. Bernard Moret\\

  \medskip\bigskip
  \Large\bf Homework Assignment \#6\\

  \smallskip
  \normalsize\rm due Sunday night, Nov.\ 4
\end{center}

\bigskip\rm\noindent
\textit{Write your solutions in LaTeX using the template provided on the
Moodle and web sites and upload your PDF file through Moodle by 4:00 am
Monday morning, Nov.\ 5.}


\bigskip\rm\noindent
\emph{Question 1.~(Matching)}\\
Consider the following simple model for mobile phones.
We have $n$ base stations and $n$ phones, all of which are specified as points
in the plane.  The phones are said to be \emph{fully connected} if each phone
can be assigned to a distinct base station, and the straight-line distance
between the assigned pair of phone and station is no more than a given
constant $c$.

Suppose the user of the first phone drives from its original point along a line
for $k$ units of distance.  As this phone moves, we have to update the
assignment~(possible several times) to keep all phones fully connected.  Give
an $O(n^3)$ running time algorithm to decide whether it is possible to keep all
phones fully connected at all times during the driving.  (Hint: a straight line
cuts a circle in at most two points.)


\bigskip\rm\noindent
\emph{Question 2.~(Stable Matching)}\\
Prove that a stable matching is both man-optimal and woman-optimal if and
only if it is the unique stable matching for the problem.

\bigskip\rm\noindent
\emph{Question 3.~(Stable Matching)}\\
Consider the stable matching problem in the case where ties are allowed in
the preference lists.
\begin{enumerate}
  \itemsep 0pt
  \item
    A \emph{strong instability} in a perfect matching $S$ consists of a man $m$
    and a woman $w$ such that $m$ and $w$ prefer each other to their partners
    in $S$.  Does there always exist a perfect matching with no strong
    instability?
    
    Give a polynomial-time algorithm that is guaranteed to find a
    perfect matching without strong instability and prove the correctness of
    your algorithm, or give an instance for which every perfect matching has a
    strong instability.
  \item
    A \emph{weak instability} in a perfect matching $S$ consists of a man $m$
    and a woman $w$ such that $m$ prefers $w$ to his partner in $S$ and $w$
    either prefers $m$ to his partner in $S$ or likes the two men equally; or
    $w$ prefers $m$ to her partner in $S$ and $m$ either prefers $w$ to her
    partner in $S$ or is likes them equally.  Does there always exist a
    perfect matching with no weak instability?
    
    Give a polynomial-time
    algorithm that is guaranteed to find a perfect matching without weak
    instability, or give an instance for which every perfect matching has a
    weak instability.
\end{enumerate}

\bigskip\rm\noindent
\emph{Question 4.~(Network Flow)}\\
We are given a directed network $G=(V,E)$ with a single specified \emph{good}
node $g\in V$ and a set of \emph{bad} nodes $B\subset V$~(naturally, we
have $g\not\in B$).  We want to disconnect bad nodes from $g$ by removing
edges, but face a tradeoff: we want to remove as few edges as possible and
yet we want to remove as many bad nodes as possible.

In consequence, we want to maximize the objective function $|f(S)| -
\alpha\cdot|S|$, where $S$ is the subset of $E$ to remove, $f(S)$ is the set of
bad nodes that cannot be reached from $g$ in the subgraph $(V, E-S)$, and
$\alpha$ is a positive constant.  Give a polynomial-time algorithm to find a
subset $S\subset E$ to maximize this objective function, and prove the
correctness of your algorithm.
\end{document}
